When Forgiveness Backfires: Rejection Sensitivity and Cooperative Behavior Following Exposure to Adaptive Forgiving Agents

### 2024-07-09 article.Rmd
version 3 finalised

---
title: "When Forgiveness Backfires: an AI-Mediated Randomised Controlled Experiment on Rejection Sensitivity and Cooperation"
author:
  # "Ismail Guennouni, Georgia Koppe, Christoph Korn"

keywords:
- Social Trust
- Repair of Cooperation
- Cognitive Manipulation 
- Economic Games 
- Hidden Markov Models

abstract: |
 Can exposure to forgiving behavior improve cooperative outcomes for individuals prone to rejection sensitivity? This randomized controlled experiment explores this question using a novel approach with Hidden Markov Model (HMM)-based AI agents in repeated trust games (RTGs). Participants (N = 206), pre-screened for high or low rejection sensitivity (RS), interacted with either consistently behaving or more forgiving co-players, which were in reality HMM agents. Unexpectedly, exposure to forgiving agents led to reduced cooperation in subsequent interactions. Control group participants maintained consistent behavior. RS levels influenced perceptions of the agent's cooperativeness but did not significantly affect behavioral outcomes, revealing a perception-behavior dissociation. These findings challenge assumptions about fostering cooperation through simple exposure to positive interactions. They also demonstrate the potential of HMM-based AI agents for studying interactive social dynamics, offering a methodological advancement for future research. Our results suggest the need for nuanced interventions that gradually shape expectations and behaviors to foster cooperative outcomes, particularly for individuals with high rejection sensitivity.

bibliography: "bib/exposure.bib"
biblio-style: spphys

output:
  bookdown::pdf_document2:
    toc: false
header-includes:
  - \usepackage{lscape}
---

@@

One explanation for such social challenges lies in early caregiver experiences. Attachment theory [@bowlby_attachment_1978] suggests that the quality of these relationships shapes our capacity for secure attachments and trust. Individuals with higher levels of insecure attachment may recall negative trust-related experiences more easily, report fewer positive trust experiences, and use less constructive coping strategies when trust is broken [@mikulincer_attachment_1998]. These insecure attachment patterns are often associated with heightened rejection sensitivity (RS), a tendency to anxiously expect, readily perceive, and intensely react to rejection [@downey_implications_1996; @downey_early_1997]. RS has been linked to the development of various mental health conditions, including depression, anxiety, personality disorders, and self-harm [@gao_associations_2017]. A recent meta-analysis revealed prosocial behavior and interpersonal trust as two key processes of interpersonal functioning that are markedly impaired in PDs and which are likely to contribute to interpersonal dysfunction in this population [@hepp_prosociality_2022]. The interaction of rejection sensitivity and trust-based constructs has been explored, particularly in Borderline Personality Disorder (BPD). @miano_rejection_2013 and @richetin_emotional_2018 found that rejection sensitivity mediates the relationship between BPD features and lower trust appraisal. However, the interaction between trustworthiness and rejection sensitivity hasn't been studied as extensively in other personality disorders or in non-clinical populations, leaving a gap in our understanding of how these constructs might interplay. Could an ingrained sensitivity to rejection contribute to reduced pro-social behavior such as a failure to show trustworthiness?

@@

A total of 206 participants (56% female) were recruited on the Prolific Academic platform (prolific.co). The required sample size was determined based on an a priori power analysis. The mean age of participants was 34.6 years, with an 11.9 years standard deviation. Participants were paid a fixed fee of £6 plus a bonus payment dependent on their performance that averaged £0.5. A total of 1195 participants were pre-screened on Prolific using the Rejection Sensitivity Questionnaire (RSQ) to finally select two similarly sized groups: One with high RS (RSQ score > 15) and the other with low RS (RSQ score < 10).

@@

The experiment had a 2 (Condition: Manipulation or Control) by 2 (RS : high or low) by 2 (Game: Trust-Game Pre Manipulation, Trust-Game Post Manipulation) design, with repeated measures on the third factor (Figure \@ref(fig:HMMPanels).A). Participants within each pre-screened group were randomly assigned to one of the two levels of the Condition factor. The games were designed and implemented online using Empirica v1 [@almaatouq_empirica_2021]. The planned experiment received approval from the University of Heidelberg's Medical Faculty ethics commission (ID:S-708/2023) and the experiment was performed in accordance with the ethics board guidelines and regulations. All participants provided informed consent prior to their participation.

@@

Participants played a 15-round RTG [@joyce_trust_1995] in the trustee role against a computer-programmed investor. On each round the investor is endowed with 20 units and decides how much of that endowment to invest. This investment is tripled and the trustee then decides how to split this tripled amount between them and the investor. If the trustee returns more than one third of the amount, the investor makes a gain. Each player was represented with an icon with the participant always on the left of the screen and the co-player on the right. The participants were able to choose the icon that represents them at the start of the experiment. The icon representing the co-player changed at the start of each new game, to simulate a new interaction partner. Participants were not told they were facing computerised co-players.

@@

At the beginning of each game in all three phases, participants were told they would face a new player and had to wait to be paired with an available co-player. This simulated the waiting time in real social interaction tasks. After completing each RTG in each phase, participants rated how cooperative and forgiving they perceived the co-playerr to be, and whether they would like to play with them again (all on a scale from 1 to 10 with 10 being the most positive rating). After completing the three game phases, participants then completed the Levels of Personality Functioning Scale (LPFS) questionnaire (see the supplement for details). Finally, participants were asked whether they thought the other players were human or computer agents, then debriefed and thanked for their participation.

@@

summary(mod_invs_RS)

@@

We used a randomized controlled online experiment where participants played a RTG with AI agents designed to simulate human-like trust-building scenarios. Participants were then exposed to either forgiving or human-like AI agents before playing another RTG. Contrary to our hypothesis, exposure to forgiving agents did not increase trust or cooperation. Instead, participants reduced their returns overall whilst the returns of those in the Control group did not change between the pre and post phase of the experiment. Neither did participants show more forgiving behavior in the Manipulation condition after the one-off defection by the agent.  Why did participants reduce their returns even though they were repeatedly exposed to a more cooperative and more forgiving AI? A look at how the participants rated their co-players might shed some light on what might be driving this reduction in returns for those in the Manipulation condition.

Those exposed to the forgiving agent rated their opponent in the post-exposure phase lower on all attributes even though they faced the same dynamic human-like HMM as pre-exposure. One possible explanation for this drop in rating is that participants exhibited a negative contrast effect. This occurs when the evaluation of a person, object, or situation is influenced by comparisons with recently encountered contrasting objects or people. If we've recently interacted with someone exceptionally nice, our perception of a normal level of niceness might be skewed, making normal behavior seem less favourable or even negative by comparison [@kobre_negative_1972]. As the most recently faced opponents were more forgiving and cooperative, this negative contrast effect may have trumped any learning transfer from being repeatedly exposed to cooperative and forgiving AI [@zentall_within-trial_2005]. If this contrast effect is indeed replicable, then an avenue for future research would be to use it to our benefit by making the participants play agents with low cooperation perception.

It is worth highlighting that RS was not correlated to overall performance in the task.  Neither did it moderate the change in returns between Conditions or have an effect on participant returns. However, in examining the player ratings by RS group more closely, we observe that individuals with high RS demonstrate a heightened attunement to changes in the behavior of their AI co-players: They accurately perceived increased cooperativeness in the forgiving AI agent, showing an ability to adjust their judgments based on interaction patterns. This may indicate that perceived support from cooperative partners can mitigate their rejection concerns. However, their declining ratings of similar co-players over time, despite consistent AI behavior, suggests that occasional pre-programmed defections may have amplified their perceptions of untrustworthiness and reluctance for future interactions. These nuanced perceptions align with the heightened social awareness often associated with high rejection sensitivity [@downey_implications_1996]. Individuals with low RS were less affected by AI behavioral variations, showing a less volatile baseline of social perception. They still exhibit behavior consistent with a contrast effect in the Manipulation condition. This stability in perception might reflect a lower propensity to scrutinize social cues or a more robust internal model of social interactions that is less easily perturbed by short-term variations in partner behavior.

@@

These findings suggest that interventions focused solely on altering behavior might be insufficient, as the underlying perceptual and cognitive processes may remain unchanged. Conversely, therapies that target these perceptual processes might not necessarily translate into immediate behavioral changes. For future research, it would be valuable to explore whether this dissociation between perception and behavior persists in more naturalistic social interactions or over longer periods. Additionally, investigating whether there are specific conditions or thresholds under which perceptual differences do begin to manifest in behavioral divergences could provide important insights into the dynamics of rejection sensitivity. These findings also suggest that simply exposing individuals to highly cooperative or forgiving partners may not be sufficient to induce lasting changes in behavior. Instead, interventions may need to incorporate more nuanced approaches that gradually shape expectations and behaviors over time, perhaps by combining periods of forgiveness with clear communication about mutual expectations for cooperation [@balliet_reward_2011]. Future research could explore the temporal dynamics of these effects, investigating whether longer exposure periods or different patterns of forgiving behavior might lead to more positive outcomes. Additionally, examining the role of explicit communication and meta-cognitive reflection on the cooperative process could provide insights into how to more effectively leverage positive interactions to promote lasting cooperative tendencies.

@@

This randomised controlled experiment enabled us to uncover unexpected effects of exposure to forgiving behavior on subsequent cooperation, particularly in relation to rejection sensitivity. These findings challenge existing assumptions about fostering cooperative behavior and suggest the need for more nuanced interventions. Importantly, the use of HMM-based AI agents in this study represents a significant methodological advancement. By providing a balance between experimental control and realistic, adaptive behavior, these agents allowed for a nuanced exploration of trust dynamics that would be challenging to achieve with human confederates or simplistic computer algorithms. This approach opens up new possibilities for studying complex social interactions in controlled settings, potentially bridging the gap between laboratory experiments and real-world social dynamics.

@@

>>>>

---
title: "When Forgiveness Backfires: Rejection Sensitivity and Cooperative Behavior Following Exposure to Adaptive Forgiving Agents"

author:
  "Ismail Guennouni^1^*, Georgia Koppe^2^, Christoph Korn^3^"

abstract: |
  Can exposure to forgiving behavior improve cooperative outcomes, and does Rejection Sensitivity (RS) moderate this effect?  This randomized controlled experiment explores these questions using a novel approach with Hidden Markov Model (HMM)-based artificial agents in repeated trust games (RTGs). Participants (N = 206), pre-screened for high or low RS, interacted with either consistently behaving or more forgiving co-players, which were in reality HMM agents. Unexpectedly, exposure to forgiving agents led to reduced cooperation in subsequent interactions. Control group participants maintained consistent behavior. RS levels influenced perceptions of the agent's cooperativeness but did not significantly affect behavioral outcomes, revealing a perception-behavior dissociation. These findings challenge assumptions about fostering cooperation through simple exposure to positive interactions and highlight the importance of accounting for negative contrast effects when designing such interventions. They also demonstrate the potential of HMM-based artificial agents for studying interactive social dynamics, offering a methodological advancement for future research. Our results suggest the need for nuanced interventions that gradually shape expectations and behaviors to foster cooperative outcomes, particularly for individuals with high rejection sensitivity.

bibliography: "bib/exposure.bib"
biblio-style: spphys
output:
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
header-includes:
  - \usepackage{lscape}
---
\small
^1^ *Central Institute of Mental Health, Medical Faculty Mannheim/Heidelberg University, Germany*

^2^ *Faculty of Mathematics and Computer Science, Heidelberg University, Germany*

^3^ *Department of General Psychiatry, Section Social Neuroscience, Heidelberg University, Germany*

^\*^ *Corresponding author. Address: Central institute of Mental Health, J5, Mannheim, Germany. Email: ismail.guennouni@zi-mannheim.de* <br>

\vspace{1cm}

**Keywords** : Social Trust; Forgiveness Intervention; Rejection Sensitivity; Interpersonal Dynamics; Hidden Markov Models

\vspace{1cm}

**Highlights:**

- Exposure to forgiving artificial agents unexpectedly led to reduced cooperation in subsequent interactions, challenging assumptions about fostering trust through positive experiences

- Rejection sensitivity influenced perceptions of artificial agents' cooperativeness but not behavioral outcomes, revealing a dissociation between perception and behavior

- Novel Hidden Markov Model-based artificial agents provide adaptive, human-like behavior in trust games, allowing for precise control of "cooperative" and "forgiving" strategies while maintaining experimental control of co-player strategies

@@

One explanation for such social challenges lies in early caregiver experiences. Attachment theory [@bowlby_attachment_1978] suggests that the quality of these relationships shapes our capacity for secure attachments and trust. Individuals with higher levels of insecure attachment may recall negative trust-related experiences more easily, report fewer positive trust experiences, and use less constructive coping strategies when trust is broken [@mikulincer_attachment_1998]. These insecure attachment patterns are often associated with heightened rejection sensitivity (RS), a tendency to anxiously expect, readily perceive, and intensely react to rejection [@downey_implications_1996; @downey_early_1997]. RS has been linked to the development of various mental health conditions, including depression, anxiety, personality disorders, and self-harm [@gao_associations_2017]. A recent meta-analysis revealed prosocial behavior and interpersonal trust as two key processes of interpersonal functioning that are markedly impaired in PDs and which are likely to contribute to interpersonal dysfunction in this population [@hepp_prosociality_2022]. The interaction of rejection sensitivity and trust-based constructs has been explored, particularly in Borderline Personality Disorder (BPD). @miano_rejection_2013 and @richetin_emotional_2018 found that rejection sensitivity mediated the relationship between BPD features and lower trust appraisal. @abramov_influence_2022-1 found that higher baseline feelings of rejection in individuals with BPD predict slower trust formation and less pronounced declines in trust following trust violations during the trust game. However, the interaction between trustworthiness and rejection sensitivity hasn't been studied as extensively in other personality disorders or in non-clinical populations, leaving a gap in our understanding of how these constructs might interplay. Could an ingrained sensitivity to rejection contribute to reduced pro-social behavior such as a failure to show trustworthiness?

@@

To have participants with large differences in RS, a total of 1195 participants were pre-screened on the Prolific Academic platform (prolific.co) using the Rejection Sensitivity Questionnaire (RSQ) to finally select two similarly sized groups: One with high RS (RSQ score > 15) and the other with low RS (RSQ score < 10) totalling 206 participants (56% female). These were then invited through prolific to take part in the main experiment study. The required sample size was determined using an a priori power analysis to have an 80% probability to detect a small effect size with a 5% type I error rate in a repeated measures ANOVA with between factors. The mean age of participants was 34.6 years, with an 11.9 years standard deviation. Participants were paid a fixed fee of £6 plus a bonus payment dependent on their performance that averaged £0.5.

@@

The experiment had a 2 (Condition: Manipulation or Control) by 2 (RS : High or Low) by 2 (Game: Trust-Game Pre Manipulation, Trust-Game Post Manipulation) design, with repeated measures on the third factor (Figure \@ref(fig:HMMPanels).A). Participants within each pre-screened group were randomly assigned to one of the two levels of the Condition factor. The games were designed and implemented online using Empirica v1 [@almaatouq_empirica_2021]. The planned experiment received approval from the University of Heidelberg's Medical Faculty ethics commission (ID:S-708/2023) and the experiment was performed in accordance with the ethics board guidelines and regulations. All participants provided informed consent prior to their participation.

@@

Participants played a 15-round RTG [@joyce_trust_1995] in the trustee role against a computer-programmed investor. On each round the investor is endowed with 20 units and decides how much of that endowment to invest. This investment is tripled and the trustee then decides how to split this tripled amount between them and the investor. If the trustee returns more than one third of the amount, the investor makes a gain. Each player was represented with an icon with the participant always on the left of the screen and the co-player on the right. The participants were able to choose the icon that represents them at the start of the experiment. The icon representing the co-player changed at the start of each new game, to simulate a new interaction partner. Participants were not told they were facing computerised co-players. We chose to simulate the behavior of a human interaction partner through allowing for a delay whilst pairing with new opponents as the start of each game as well as programming the agents to respond during each round after a varying time lapse (randomly chosen between 5 and 10 seconds).

@@

At the beginning of each game in all three phases, participants were told they would face a new player and had to wait to be paired with an available co-player. This simulated the waiting time in real social interaction tasks. After completing each RTG in each phase, participants rated how cooperative and forgiving they perceived the co-playerr to be, and whether they would like to play with them again (all on a scale from 1 to 10 with 10 being the most positive rating). After completing the three game phases, participants then completed the Levels of Personality Functioning Scale Brief-Form (LPFS-BF) questionnaire [@weekers_level_2019]. This is a self-report measure designed to assess core elements of personality functioning as defined in the Alternative Model for Personality Disorders in the DSM-5 [@american_psychiatric_association_diagnostic_2013], and provides a dimensional assessment of personality functioning, which complements the categorical approach of RS. Finally, participants were asked whether they thought the other players were human or computer agents, then debriefed and thanked for their participation.

@@

summary(mod_invs_RS)
```

```{r}
saveRDS(summary(mod_invs_RS), "data/mod_invs_RS.RDS")
```

@@

We used a randomized controlled online experiment where participants played a RTG with artificial agents designed to simulate human-like trust-building scenarios. Participants were then exposed to either forgiving or human-like HMM agents before playing another RTG. Contrary to our hypothesis, exposure to forgiving agents did not increase trust or cooperation. Instead, participants reduced their returns overall whilst the returns of those in the Control group did not change between the pre and post phase of the experiment. Neither did participants show more forgiving behavior in the Manipulation condition after the one-off defection by the agent.  Why did participants reduce their returns even though they were repeatedly exposed to more cooperative and more forgiving artificial agents? A look at how the participants rated their co-players might shed some light on what might be driving this reduction in returns for those in the Manipulation condition.

Those exposed to the forgiving agent rated their opponent in the post-exposure phase lower on all attributes even though they faced the same dynamic human-like HMM as pre-exposure. One possible explanation for this drop in rating is that participants exhibited a negative contrast effect. This occurs when the evaluation of a person, object, or situation is influenced by comparisons with recently encountered contrasting objects or people. If we've recently interacted with someone exceptionally nice, our perception of a normal level of niceness might be skewed, making normal behavior seem less favourable or even negative by comparison [@kobre_negative_1972]. As the most recently faced opponents were more forgiving and cooperative, this negative contrast effect may have trumped any learning transfer from being repeatedly exposed to cooperative and forgiving agents [@zentall_within-trial_2005]. If this contrast effect is indeed replicable, then an avenue for future research would be to use it to our benefit by making the participants play agents with low cooperation perception.

It is worth highlighting that RS was not correlated to overall performance in the task.  Neither did it moderate the change in returns between Conditions or have an effect on participant returns. However, in examining the player ratings by RS group more closely, we observe that individuals with high RS demonstrate a heightened attunement to changes in the behavior of their HMM co-players: They accurately perceived increased cooperativeness in the forgiving agent, showing an ability to adjust their judgments based on interaction patterns. This may indicate that perceived support from cooperative partners can mitigate their rejection concerns. However, their declining ratings of similar co-players over time, despite consistent agent behavior, suggests that occasional pre-programmed defections may have amplified their perceptions of untrustworthiness and reluctance for future interactions. These nuanced perceptions align with the heightened social awareness often associated with high rejection sensitivity [@downey_implications_1996]. Individuals with low RS were less affected by the agent's behavioral variations, showing a less volatile baseline of social perception. They still exhibit behavior consistent with a contrast effect in the Manipulation condition. This stability in perception might reflect a lower propensity to scrutinize social cues or a more robust internal model of social interactions that is less easily perturbed by short-term variations in partner behavior.

@@

These findings suggest that interventions focused solely on altering behavior might be insufficient, as the underlying perceptual and cognitive processes may remain unchanged. Conversely, therapies targeting these perceptual processes might not necessarily translate into immediate behavioral changes. Future research should investigate whether this dissociation between perception and behavior persists in more naturalistic social interactions or over longer periods. Additionally, exploring specific conditions or thresholds under which perceptual differences begin to manifest in behavioral divergences could provide valuable insights into the dynamics of rejection sensitivity. Simply exposing individuals to highly cooperative or forgiving partners may not suffice to induce lasting changes in behavior. Instead, interventions may need to incorporate more nuanced approaches that gradually shape expectations and behaviors over time. Combining periods of forgiveness with clear communication about mutual expectations for cooperation might be more effective [@balliet_reward_2011]. Future studies should also consider the temporal dynamics of these effects, examining whether longer exposure periods or different patterns of forgiving behavior could lead to more positive outcomes. Moreover, investigating the role of explicit communication and meta-cognitive reflection on the cooperative process could offer insights into leveraging positive interactions to promote enduring cooperative tendencies.

@@

This randomised controlled experiment enabled us to uncover unexpected effects of exposure to forgiving behavior on subsequent cooperation, particularly in relation to rejection sensitivity. These findings challenge existing assumptions about fostering cooperative behavior and suggest the need for more nuanced interventions. Importantly, the use of HMM-based artificial agents in this study represents a significant methodological advancement. By providing a balance between experimental control and realistic, adaptive behavior, these agents allowed for a nuanced exploration of trust dynamics that would be challenging to achieve with human confederates or simplistic computer algorithms. This approach opens up new possibilities for studying complex social interactions in controlled settings, potentially bridging the gap between laboratory experiments and real-world social dynamics.

@@

# Author contributions statement {.unnumbered}
I. Guennouni, G. Koppe and C. Korn. designed and developed the study concept. Experiment design, testing and data collection were performed by I. Guennouni. I. Guennouni analysed and interpreted the data under the supervision of G. Koppe and C. Korn. All authors jointly wrote and approved the final version of the manuscript for submission.

# Funding {.unnumbered}
This publication was supported through state funds approved by the State Parliament of Baden-Württemberg for the Innovation Campus Health + Life Science alliance Heidelberg Mannheim. I. Guennouni was supported by a research fellowship from the AI Health Innovation Cluster

# Competing interests statement {.unnumbered}
The author(s) declared that there were no conflicts of interest with respect to the authorship or the publication of this article.

# Additional information  {.unnumbered}

## Correspondence {.unnumbered}

All correspondence and requests for materials should be addressed to I. Guennouni.

## Transparency and data availability {.unnumbered}

Preregistration: The hypotheses and methods were not preregistered.The analysis plan was not preregistered. Materials: All study materials are publicly available (https://github.com/ismailg/exposure-public). Data: All primary data are publicly available (https://github.com/ismailg/exposure-public). Analysis scripts: All analysis scripts are publicly available (https://github.com/ismailg/exposure-public).
